How Kids Code and How We Know:\\An Exploratory Study on the Scratch Repository

### 2016-04-22 Paper/paper.tex
Changes based on Felienne's comments

Our study is based on data from the Scratch project repository. By scraping the list of recent projects,\footnote{\label{scratchpublic}\url{https://scratch.mit.edu/explore/projects/all/}} we have obtained \nPrograms~public Scratch projects and performed analysis on them. To the best of our knowledge, this is the first large-scale exploratory study of Scratch projects.

@@

\begin{itemize}
	\item{A public data set of \nScriptPrograms~non-empty Scratch projects (Section \ref{dataset})}
	\item{An evaluation of the data set in terms of size, complexity, utilized programming concepts and features, and identified smells (Section \ref{sec:results})}
	\item{A discussion of the implications of our findings for educational programming language designers (Section \ref{sec:discussion})}
\end{itemize}

@@

\begin{description}
	\item [Sprites] Scratch code is organized by `sprites': two dimensional pictures each having their own associated code. Scratch allows users to bring their sprites to life in various ways, for example by moving them in the plane, having them say or think words or sentences via text balloons, but also by having them make sounds, grow, shrink and switch costumes. The Scratch program in Figure \ref{fig:ui} consists two one sprites, the cat, which is Scratch's default sprite and logo and a piano. The code in Sprite1 will cause the cat to move right when the right arrow is pressed, and when the green flag is clicked it will continuously sense touching the piano.
	\item [Scripts]Sprites can have multiple code blocks, called scripts. The Scratch code in Figure \ref{fig:ui} has two distinct scripts, one started by clicking on the green flag and one by pressing the space bar. It is possible for a single sprite to have multiple scripts initiated by the same event. In that case, all scripts will be executed simultaneously.
	\item [Events] Scratch is \emph{event-driven}: all motions, sounds and changes in the looks of sprites are initiated by events called Hat blocks\footnote{\url{http://wiki.scratch.mit.edu/wiki/Hat_Block}}). The canonical event is the \texttt{when Green Flag clicked}, activated by clicking the green flag at the top of the user interface. In addition to the green flag, there are a number of other events possible, including key presses, mouse clicks and input from a computer's microphone or webcam. The Scratch code in Sprite1 in Figure \ref{fig:ui} contains two events: \texttt{when Green Flag clicked} and \texttt{when right arrow key pressed}, each with associated blocks.
	\item [Signals] Events within Scratch can be user generated too: users can broadcast a message, for example when two sprites touch each other, like in Figure \ref{fig:ui}. All other sprites can then react by using the \texttt{when I receive} Hat block. In Figure \ref{fig:ui}, Sprite1 broadcasts `bump' when the cat touches the Piano. 
	\item[Custom blocks] Scratch users can define their own blocks, which users can name themselves, called custom blocks. The creation of custom blocks in Scratch is the equivalent of defining procedures in other languages \cite{moreno_automatic_2014}. Because the term `procedures' is common in the software engineering area, for readability we will refer to custom blocks as `procedures' in the following. Procedures can have input parameters of type string, number, and boolean. When a user defines a procedure, a new Hat block called \texttt{define} appears, which users can fill with the implementation of their block.
\end{description}

@@

We obtained the set of Scratch projects by using a scraping program. Our scraping program called Kragle starts by reading the Scratch projects page\footnotemark[\ref{scratchpublic}] and thus obtaining project ids of projects that were most recently shared. Subsequently, Kragle obtains the JSON code for each of the listed projects.

@@

We ran Kragle on March 2nd 2016 for 24 hours, when it had obtained a little over 250,000 projects. Out of the \nPrograms~, we failed to parse and further analyze 2,367 projects due technical difficulties with the provided JSON files. Kragle, as well as all scraped projects and our analysis files are available.\footnote{\label{repo}\url{https://github.com/ScratchLover42/ICER-Data-Code}}

Once we obtained the Scratch projects, we parsed the JSON files according to the specification of the format.\footnote{\url{http://wiki.scratch.mit.edu/wiki/Scratch_File_Format_(2.0)}} This resulted in a list of used blocks per project, with the sprites and the stage of the project. We cross referenced also all blocks with the Scratch wiki so we have the shapes and the category of all blocks. For example, \texttt{When Green Flag Clicked} is a \emph{Hat block} from the \emph{Events category}.

\subsection{Data analysis}
\label{dataAnalysis}
All scraped project data, including the list of used blocks and parameters, was imported in a relational database. We used SQL queries, which are also made available,\footnotemark[\ref{repo}] for filtering, aggregating and extracting all statistical data required to address our three research questions. We also randomly sampled and manually inspected edge cases in the results, for example empty or overly complex projects, lists of which we also provide as part of the dataset.\footnotemark[\ref{repo}]

For RQ1 we measured the size of projects based on the number of blocks in scripts and sprites and we calculated descriptive statistics, which are presented in Section \ref{RQ1}. We used the McCabe cyclomatic complexity metric\cite{mcCabe76} for measuring the complexity of the scripts.

For RQ2, we used the data on the code blocks and their categories to perform a detailed statistical analysis of applied programming abstractions and concepts. Similarly to \cite{Maloney_2008}, we consider the use of certain blocks to indicate that a programming abstraction or concept is being used in a certain project. In Section \ref{RQ2} we present the results related to the utilization of procedures, variables, loops, conditional statements, user interactivity and synchronization.

For RQ3, we focused on four types of code smells: duplicated code, dead code, large script and large sprite. For the duplicated code smell analysis, our first step was to specify what we consider a code clone in the context of Scratch programming: a script that is composed of a set of blocks of the same type connected in the same way and is repeated within or across sprites of the same project. For the identification of clones we did not take into account the values of the parameters that may be used in the blocks, so that two blocks that only differ in the values of parameters are considered to be equal. We also examined the case of clones with the same parameter values, and we refer to them as \textit{exact clones}. The next step in the analysis was to determine the minimum size of the scripts that are considered clones instead of incidentally similar. For this, we examined the number of detected clones for different script sizes and present the results in Figure \ref{fig:cloneslines}. Based on this distribution, we opted to adopt the number also used by the authors in \cite{moreno_automatic_2014}, which is the minimum size of 5 blocks per script.

@@

To examine the long method and the large class smells, we consider them in the context of Scratch as large script and large sprite smells respectively. For these two smells we use the number of blocks as the size metric. Figure \ref{fig:longmethod} presents the number of blocks in the scripts and the sprites of our dataset. We used these numbers to split the dataset and retrieve the top 10\% largest scripts and sprites, as is commonly done in both source code analysis~\cite{alves_deriving_2010} and analysis of end-user programming artifacts like spreadsheets~\cite{hermans2015detecting}. Using that strategy, we set the thresholds to be used for the calculation of the large script and large sprite: it is 18 blocks and 59 blocks respectively. The results we obtained using these thresholds are presented in Section \ref{RQ3}.

@@

In Table \ref{tbl-size} we summarize statistics for the analyzed metrics. We use the mean value and the five-number summary to describe the dataset in terms of the number of sprites with code per project (including the stage sprite) and the number of scripts and blocks per project. Figures \ref{fig:size-sprites}, \ref{fig:size-scripts}, \ref{fig:size-locs} and plot the distribution of these size metrics.

@@

To understand the complexity of the Scratch projects in our dataset, we use the McCabe cyclomatic complexity \cite{mcCabe76}, a quantitative measure of the number of independent paths through a program's source code. This is calculated per script by counting the number of decision points in the script plus one. In Scratch, decision points can be the \texttt{if} and \texttt{if else} blocks. The results of the cyclomatic complexity metric per script are plotted in Figure \ref{fig:cyclomatic}. The majority (78.33\%) of \nscripts~scripts contain no decision points, while 13.08\% have a cyclomatic complexity of 2, containing exactly one decision point. The complexity is higher, over 4, for 3.67\% of the scripts. The analysis also highlighted 209 scripts with a cyclomatic complexity over 100 and up to 246.\footnotemark[\ref{repo}] Cyclomatic complexity was greater (mean value of 3.32\%) in defined procedures, with 56.46\% of the procedures having at least one decision point.

@@

As shown in Table \ref{tbl-characteristics}, almost one-third of the projects use variables and a small number (4.01\%) use lists. The number of variables that is being used is also limited, with only 7.48\% of the projects having 5 or more variables. The distribution of variable and list utilization is shown in Figure \ref{fig:variables}. Exceptional cases exist: the analysis highlighted 842 projects with more than 100 variables and up to 340. Examining the initialization of variables through the \texttt{set <variable> to <value>} blocks, we found that for 4.83\% of all variables this was missing. While failing to initialize a variable in Scratch will not result in a runtime error as in some other programming languages, correctly setting the initial state of the program is important \cite{boe_hairball:_2013}.

@@

The third and fourth cases are detected in scripts that are never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, or scripts that are comprised of only a \texttt{when <trigger>} block without any functionality. A total of 322,475 scripts like that were found in 56,890 (24.36\% of the total) projects. The majority of these scripts (86.6\%) are scripts missing the starting block. Examining the size of these dead scripts, 72.34\% are composed of a single block. As shown in Figure \ref{fig:deadcode}, however, considerably large dead scripts exist; 2,358 of these scripts originating from 1,553 different projects have more than 30 blocks and up to 2,610.\footnotemark[\ref{repo}] The number of projects exhibiting the dead code smell, considering all four types of dead code, is 65,760 (28.16\% of the total projects).

@@

Finally, we examine the large script and the large sprite smells. The thresholds we use for the identification of large scripts and large sprites are 18 blocks and 59 blocks respectively, as calculated in Section \ref{dataAnalysis}. The number of projects exhibiting the large script smell, i.e., containing at least one script with 18 or more blocks, is 69,521 (29.77\% of the total projects) and the number of projects with the large sprite smell is 31,954 (13.68\%).

@@

In a sense, the dead scripts are harmless, as they are not executed. However, they do cause `visual clutter' and might be distracting to novice programmers, as it might be hard to see which scripts are dead. In contrast to other visual educational languages, Scratch does on indicate scripts that are dead. LEGO Mindstorms, for example, does give the user feedback by making unconnected blocks gray.

@@

A study on the internalization of programming concepts with Scratch with 46 students was presented in \cite{meerbaum-salant_learning_2010}. Concepts like loops, conditional loops, message passing, initialization, variables and concurency were examined, and it was found that students had problems with the last three. In a later study with an equal set of subjects \cite{Meerbaum_habits_2011} the same authors identified two bad programming habits in Scratch, namely bottom-up development and extremely fine-grained programming. They connected the later to the reduced use of if-blocks and finite loops and the increased use of infinite loops, a finding that is verified by our study. In \cite{wilson_evaluation_2012} 29 Scratch projects created from 60 students working in groups were evaluated based on a list of criteria related to programming concepts, code organization and usability design.

@@

\section{Conclusions}
\label{sec:conclusion}
In this paper we presented a large-scale study on 247,798 projects we scraped from the Scratch repository. We analyzed these projects in terms of size, complexity, application of programming abstractions and utilization of programming concepts including procedures, variables, conditional statements, loops, and broadcast-receive functionality. We found that procedures and conditional loops are not commonly used. We further investigated the presence of code smells, including code duplication, dead code, long method and large class smells. Our findings indicate that Scratch programs suffer from code smells and especially from dead code and code duplication.

>>>>

Our study is based on data from the Scratch project repository. By scraping the list of recent projects,\footnote{\label{scratchpublic}\url{https://scratch.mit.edu/explore/projects/all/}} we have obtained \nPrograms~public Scratch projects and performed source code analysis on them. To the best of our knowledge, this is the first large-scale exploratory study of Scratch projects.

@@

\begin{itemize}
	\item{A public data set of \nScriptPrograms~non-empty Scratch projects (Section \ref{dataset})}
	\item{An evaluation of the data set in terms of size, complexity, programming concepts and smells (Section \ref{sec:results})}
	\item{A discussion of the implications of our findings for educational programming language designers (Section \ref{sec:discussion})}
\end{itemize}

@@

\begin{description}
	\item [Sprites] Scratch code is organized by `sprites': two dimensional pictures each having their own associated code. Scratch allows users to bring their sprites to life in various ways, for example by moving them in the plane, having them say or think words or sentences via text balloons, but also by having them make sounds, grow, shrink and switch costumes. The Scratch program in Figure \ref{fig:ui} consists two one sprites, the cat, which is Scratch's default sprite and logo and a piano. The code in Sprite1 will cause the cat to move right when the right arrow is pressed, and when the green flag is clicked it will continuously sense touching the piano.
	\item [Scripts]Sprites can have multiple code blocks, called scripts. The Scratch code in Figure \ref{fig:ui} has two distinct scripts, one started by clicking on the green flag and one by pressing the space bar. It is possible for a single sprite to have multiple scripts initiated by the same event. In that case, all scripts will be executed simultaneously.
	\item [Events] Scratch is \emph{event-driven}: all motions, sounds and changes in the looks of sprites are initiated by events called Hat blocks\footnote{\url{http://wiki.scratch.mit.edu/wiki/Hat_Block}}). The canonical event is the \texttt{when Green Flag clicked}, activated by clicking the green flag at the top of the user interface. In addition to the green flag, there are a number of other events possible, including key presses, mouse clicks and input from a computer's microphone or webcam. The Scratch code in Sprite1 in Figure \ref{fig:ui} contains two events: \texttt{when Green Flag clicked} and \texttt{when <right arrow> key pressed}, each with associated blocks.
	\item [Signals] Events within Scratch can be user generated too: users can broadcast a message, for example when two sprites touch each other, like in Figure \ref{fig:ui}. All other sprites can then react by using the \texttt{when I receive} Hat block. In Figure \ref{fig:ui}, Sprite1 broadcasts `bump' when the cat touches the Piano. 
	\item[Custom blocks] Scratch users can define their own blocks, which users can name themselves, called custom blocks. The creation of custom blocks in Scratch is the equivalent of defining procedures in other languages \cite{moreno_automatic_2014}. Because the term `procedures' is common in the software engineering area, for readability we will refer to custom blocks as `procedures' in the following. Procedures can have input parameters of type string, number, and boolean. When a user defines a procedure, a new Hat block called \texttt{define} appears, which users can fill with the implementation of their block.
\end{description}

@@

We obtained the set of Scratch projects by using a scraping program. Our scraping program called Kragle starts by reading the Scratch projects page\footnotemark[\ref{scratchpublic}] and thus obtains project ids of projects that were most recently shared. Subsequently, Kragle retrieves the JSON code for each of the listed projects.

@@

We ran Kragle on March 2nd 2016 for 24 hours, when it had obtained a little over 250,000 projects. Out of the \nPrograms, we failed to parse and further analyze 2,367 projects due technical difficulties with the provided JSON files. Kragle, as well as all scraped projects and our analysis files are available.\footnote{\label{repo}\url{https://github.com/ScratchLover42/ICER-Data-Code}}

Once we obtained the Scratch projects, we parsed the JSON files according to the specification of the format.\footnote{\url{http://wiki.scratch.mit.edu/wiki/Scratch_File_Format_(2.0)}} This resulted in a list of used blocks per project, with the sprites and the stage of the project. We cross referenced also all blocks with the Scratch wiki to determine the shapes and the category of all blocks. For example, \texttt{When Green Flag Clicked} is a \emph{Hat block} from the \emph{Events category}.

\subsection{Data analysis}
\label{dataAnalysis}
All scraped project data, including the list of used blocks and parameters, were imported in a relational database. We used SQL queries, which are also made available,\footnotemark[\ref{repo}] for filtering, aggregating and extracting all statistical data required to address our three research questions. We also randomly sampled and manually inspected edge cases in the results, for example empty or overly complex projects, lists of which we also provide as part of the dataset.\footnotemark[\ref{repo}]

For RQ1 we measured the size of projects based on the number of blocks in scripts and sprites and we calculated descriptive statistics, which are presented in Section \ref{RQ1}. For measuring the complexity of the scripts we used the McCabe cyclomatic complexity metric \cite{mcCabe76}, a quantitative measure of the number of independent paths through a program's source code. This is calculated per script by counting the number of decision points in the script plus one. In Scratch, decision points can be the \texttt{if} and \texttt{if else} blocks.

For RQ2, we used the data on the code blocks and their categories to perform statistical analysis of applied programming abstractions and concepts. Similarly to \cite{Maloney_2008}, we consider the use of certain blocks to indicate that a programming abstraction or concept is being used in a certain project. In Section \ref{RQ2} we present the results related to the utilization of procedures, variables, loops, conditional statements, user interactivity and synchronization.

For RQ3, we focused on four types of code smells: duplicated code, dead code, large script and large sprite. For the duplicated code smell analysis, our first step was to specify what we consider a code clone in the context of Scratch programming: a script that is composed of a set of blocks of the same type connected in the same way and is repeated within or across sprites of the same project. For the identification of clones we did not take into account the values of the parameters that may be used in the blocks, so that two blocks that only differ in the values of parameters are considered to be equal. We also examined the case of clones with the same parameter values, and we refer to them as \textit{exact clones}. The next step in the analysis was to determine the minimum size of the scripts that are considered clones instead of incidentally similar. We examined the number of detected clones for different script sizes and present the results in Figure \ref{fig:cloneslines}. Based on this distribution, we opted to adopt the number also used by the authors in \cite{moreno_automatic_2014}, which is the minimum size of 5 blocks per script.

@@

To examine the long method and the large class smells, we consider them in the context of Scratch as large script and large sprite smells respectively. For these two smells we use the number of blocks as the size metric. Figure \ref{fig:longmethod} presents the number of blocks in the scripts and the sprites of our dataset. We used these numbers to split the dataset and retrieve the top 10\% largest scripts and sprites, as is commonly done in both source code analysis~\cite{alves_deriving_2010} and analysis of end-user programming artifacts like spreadsheets~\cite{hermans2015detecting}. Using that strategy, we set the thresholds for the calculation of the large script and large sprite: it is 18 blocks and 59 blocks respectively. The results we obtained using these thresholds are presented in Section \ref{RQ3}.

@@

In Table \ref{tbl-size} we summarize the statistics for the analyzed metrics. We use the mean value and the five-number summary to describe the dataset in terms of the number of sprites with code per project (including the stage sprite) and the number of scripts and blocks per project. Figures \ref{fig:size-sprites}, \ref{fig:size-scripts}, \ref{fig:size-locs} and plot the distribution of these size metrics.

@@

To understand the complexity of the Scratch projects in our dataset, we use the McCabe cyclomatic complexity. The results of this metric per script are plotted in Figure \ref{fig:cyclomatic}. The majority (78.33\%) of \nscripts~scripts contain no decision points, while 13.08\% have a cyclomatic complexity of 2, containing exactly one decision point. The complexity is higher, over 4, for 3.67\% of the scripts. The analysis also highlighted 209 scripts with a cyclomatic complexity over 100 and up to 246.\footnotemark[\ref{repo}] Cyclomatic complexity was greater (mean value of 3.32\%) in defined procedures, with 56.46\% of the procedures having at least one decision point.

@@

As shown in Table \ref{tbl-characteristics}, almost one-third of the projects use variables and a small number (4.01\%) use lists. The number of variables that is being used is also limited, with only 7.48\% of the projects having 5 or more variables. The distribution of variable and list utilization is shown in Figure \ref{fig:variables}. Exceptional cases exist: the analysis highlighted 842 projects with more than 100 variables and with a maximum of 340. Examining the initialization of variables through the \texttt{set <variable> to <value>} blocks, we found that for 4.83\% of all variables this was missing. While failing to initialize a variable in Scratch will not result in a runtime error as in some other programming languages, correctly setting the initial state of the program is important \cite{boe_hairball:_2013}.

@@

The third and fourth cases are incomplete scripts. They are either never invoked due to the lack of a starting \texttt{when <trigger>} block from the Scratch Events or Control category, or are comprised of only a \texttt{when <trigger>} block without any functionality. A total of 322,475 scripts like that were found in 56,890 (24.36\% of the total) projects. The majority of these scripts (86.6\%) are scripts missing the starting block. Examining the size of these dead scripts, 72.34\% are composed of a single block. As shown in Figure \ref{fig:deadcode}, however, considerably large dead scripts exist; 2,358 of these scripts originating from 1,553 different projects have more than 30 blocks and up to 2,610.\footnotemark[\ref{repo}]

The number of projects exhibiting the dead code smell, considering all four types of dead code combined, is 65,760 (28.16\% of the total projects).

@@

Finally, we examine the large script and the large sprite smells. The thresholds we use for the identification of large scripts and large sprites are 18 blocks and 59 blocks respectively, as explained in Section \ref{dataAnalysis}. The number of projects exhibiting the large script smell, i.e., containing at least one script with 18 or more blocks, is 69,521 (29.77\% of the total projects) and the number of projects with the large sprite smell is 31,954 (13.68\%).

@@

In a sense, the dead scripts are harmless, as they are not executed. However, they do cause `visual clutter' and might be distracting to novice programmers, as it might be hard to see which scripts are dead. In contrast to other visual educational languages, Scratch does not indicate scripts that are dead. LEGO Mindstorms, for example, does give the user such feedback by making unconnected blocks gray.

@@

For example, a study on the internalization of programming concepts with Scratch with 46 students was presented in \cite{meerbaum-salant_learning_2010}. Concepts like loops, conditional loops, message passing, initialization, variables and concurency were examined, and it was found that students had problems with the last three. In a later study with an equal set of subjects \cite{Meerbaum_habits_2011} the same authors identified two bad programming habits in Scratch, namely bottom-up development and extremely fine-grained programming. They connected the later to the reduced use of if-blocks and finite loops and the increased use of infinite loops, a finding that is verified by our study. In \cite{wilson_evaluation_2012} 29 Scratch projects created from 60 students working in groups were evaluated based on a list of criteria related to programming concepts, code organization and usability design.

@@

\section{Conclusions}
\label{sec:conclusion}
In this paper we presented a large-scale study on 247,798 projects we scraped from the Scratch repository. We analyze these projects in terms of size, complexity, application of programming abstractions and utilization of programming concepts including procedures, variables, conditional statements, loops, and broadcast-receive functionality. We find that procedures and conditional loops are not commonly used. We further investigate the presence of code smells, including code duplication, dead code, long method and large class smells. Our findings indicate that Scratch programs suffer from code smells and especially from dead code and code duplication.
